The binary liquid model for gravity-driven microchannel simulations

### 2011-01-05 Text/paper.tex
Fix typos + style fixes in the abstract and intro.

\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{shurik.kuzmin@gmail.com}
\author[google]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[uofa]{J.~Derksen}
\ead{kuzmin@ualberta.ca}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\address[google]{Google, Zurich, Switzerland}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[schlum]{DBR Schlumberger Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}

\begin{abstract}
This work shows that the free-energy binary liquid lattice Boltzmann method is able to
reproduce the gravity driven microchannel flow of a gas bubble through liquid. In the regime of low
capillary
numbers
$\Ca$ (defined in Eq. \ref{capillary:number:definition}), the flow is thoroughly described by
Bretherton \cite{bretherton},
with the deposition film thickness proportional to $\Ca^{2/3}$. However, there are deviations
from the Bretherton correlation for $Ca\geq 0.005$, studied experimentally and numerically. This
work aims to examine the lattice Boltzmann binary liquid model feasibility to obtain valid results
for a bubble flow in liquid for a range of $Ca$ number. The binary
liquid model is a continuous-interface multiphase model allowing to properly resolve the interface.
It was found that the interface thickness needs to be $50$
percent of the film thickness for simulation to give convergent results. While
the convergence criteria is fulfilled the results are in a good agreement with the
numerical simulations of \citet{giavedoni-numerical} in the range of $Ca$ from $0.01$ to $1$. 
\end{abstract}

@@

While there are number
of works which simulate the Bretherton problem by the boundary value approach
\cite{ingham-plates,heil-bretherton}, those methods are of limited applicability for problems
involving complex geometries, free interface motion, or for problems involving coalescence and/or
droplet breakup. The continuous interface models are more flexible for those kind of simulations.
However, if the interface is spread over several grid nodes, the question of
proper film resolution in comparison with the interface resolution arises.  This work is focused on
resolving the parameter range of the binary liquid lattice
Boltzmann method to validate it in the range of capillary numbers.

The lattice Boltzmann method has emerged as a successful method to simulate
a wide variety of phenomena including hydrodynamics \cite{yu}, thermal flows
\cite{karlin-minimalmodels}, microflows \cite{ansumali-small-knudsen},
ferrofluids \cite{kuzmin-aniso}, and multiphase flows
\cite{swift,Shan-chen:extended}. Thanks to its kinetic nature, LBM as a particle
method easily tackles complex geometries and allows incorporation of
physical phenomena on the microscopic level, as in the case of multiphase models. Most
multiphase lattice Boltzmann models \cite{swift, Shan-chen:extended} resolve
the interface using continuous interface methods where the interfaces spans over several grid nodes.
Such a representation
brings issues of the film thickness resolution versus interface
resolution. Thus, the diffusive interface should be resolved in such a way as to have
negligible effect on the physics of the film.

@@

\section{Lattice Boltzmann benchmark}
To properly simulate and compare simulation results with the data
published in the literature, one needs to design a numerical benchmark addressing
the following challenges for the lattice Boltzmann method:
\begin{itemize}
 \item The diffusive nature of the interface in the multiphase model should not
influence the film thickness results. This is obviously achieved only for certain relation of the
interface resolution related to the film thickness. The same applies to spurious
currents arising due to surface tension discretization, which
can influence mass transfer in the thin film, thereby compromising the
overall solution. Therefore, the diffusive interface should not noticeably affect the film
thickness.
 \item The wettability coefficient which in principle can control the
dynamic angle and the menisci appearance \cite{pagonabarraga-finger} should not
influence the overall solution.
 \item The free-energy model examined in this work is a single density, two
viscosities model. The classical Bretherton problem is a free-interface problem between liquid and
gas
where the film thickness is established by itself. Thus, the viscosity and the density ratios for
the
classical
Bretherton problem should
be large. However, in terms of the lattice Boltzmann binary liquid framework only the viscosity
ratio can be addressed. Our computations show that the results are consistent with other works and
inertia effects can be neglected.
  \item The Bretherton problem addresses a bubble flow driven by a pressure
difference, not a body force. However, to the best authors' knowledge the pressure boundary
conditions are still not yet developed for the binary liquid model, because it involves the
treatment of associated gradients and laplacians on the boundary.
This kind of pressure boundary conditions for the binary liquid model is currently
under development, and preliminary results show that the differences between flows driven by a body
force and by
pressure difference are small as far as the film
 thickness is of interest.  In this work we limit ourselves to the study
of body force driven flows, because of their simplicity and better numerical stability. This
case implies the periodic boundary conditions. As soon as the periodic boundary
conditions are applied, not a single bubble but whole bubble train is simulated. In this case one
needs
to ensure that the distance between bubbles is large enough to exclude mutual bubble influence.
 \item Most simulations are done for a channel with circular cross section,
which is quite difficult to address in terms of lattice Boltzmann free energy
framework, where one needs to calculate the Laplacian and gradients near boundaries.
Although one can apply certain finite difference stencils
\cite{arnold-boundary,hunt-boundary} developed specifically for curved boundaries,
the analysis of error introduced by the boundary conditions becomes an issue. Though the circular
axis-symmetric case can be simulated on rectangular grid by introducing specific mass and force
terms in the continuity and Navier-Stokes equation \cite{halliday-circular}, the lattice Boltzmann
incorporation of force and mass terms is in the process of development and validation. Therefore,
the benchmark is chosen as a plane-symmetric two-dimensional case.

@@

\subsection{The influence of the wall gradient}
The simulation results should not depend on the wall wettability. Wettability is defined
through the phase gradient \cite{pooley-contact}, $\partial_n \phi$.  We took a large
enough grid resolution in order for simulations to be consistent -- the grid size was
$177 \times 2626$ and the initial film thickness was
$12$ lattice boltzmann units, which corresponds to the predicted film thickness.
We examined $11$ different values for the wall
gradient ranging from $-1$ to $1$. The parameters are summarized in Table
\ref{table:parameters:wall:gradient}. The results are consistent for all the
wall gradients given that the interface is properly resolved. For 
given wall gradients the values of interface thickness values are of $0.2\%$ relative accuracy.  For
the given wall gradients $0.8$
and $1.0$ the
simulations are unstable. One can see the calculated profiles in Fig.
\ref{fig:gradients:profiles}. The latter shows that the negative wall gradient values cause 
non-physical phase values (above $1.0$ near the wall). One should attribute them to the numerical
impose of wall gradients through the first-order finite difference scheme. Therefore, they do
not spoil numerical results. Note that
negative values of the phase gradient are
preferable, since the phase of the liquid adjacent to the wall has the value of $1$.
The phase values near the wall are above $1$ and do not interact
with the bubble which has phase value $-1$.  In case of positive
gradients the values near the wall are below $1$, and the gradient profile can
fuse with the values of the gas phase. This case corresponds to the slug flow when gas is in
contact with the wall.
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/Wall/phase_grad_profiles.eps}
\caption{Phase profiles for different wall gradients. One can see that if the thickness
is properly resolved then the wettability of the wall does affect the film thickness. The
simulations were conducted for $200,000$ iterations. The measured profile was taken at the
nondimensional coordinate scaled to $H_{\mathrm{eff}}$ as $x=11.42$. $Ca=0.05$ was taken to
initialize simulation.
\label{fig:gradients:profiles}
}
\end{figure}

@@

\section{Conclusion}
This work presents numerical tips and a benchmark for the
Bretherton/Taylor problem using the binary liquid lattice Botlzmann method. To obtain correlations
we specifically designed the benchmark. The bubble was chosen sufficiently long for the film
thickness to
stabilize. Also, we used periodic boundary conditions for scheme robustness. Thus, the bubble
train was simulated instead of one bubble motion. To avoid the mutual influence of neighboring
bubbles a
distance between bubbles was chosen to be $3$ bubble sizes. The computational
results in terms of capillary number dependence and shape of a bubble show consistency with the
previously published data. Surprisingly, with large enough viscosity ratio the results are
independent on inertia. Also, we examined the influence of grid resolution on the results. We found
that if the phase interface is resolved as $50$ percent of the simulated film thickness, the
simulations
are convergent. Though  our results are specific to the binary liquid lattice
Botlzmann method, the numerical hints and procedures can be used for any
continuous interface method. 
\bibliographystyle{model1-num-names}
\bibliography{paper}
%\bibliographystyle{unsrtnat}
%\bibliography{paper}
\end{document}

>>>>

\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{shurik.kuzmin@gmail.com}
\author[google]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[uofa]{J.~Derksen}
\ead{kuzmin@ualberta.ca}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\address[google]{Google Switzerland GmbH, Brandschenkestrasse 110, 8002 Zurich, Switzerland}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[schlum]{DBR Schlumberger Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}

\begin{abstract}
This work shows that the free-energy binary liquid lattice Boltzmann method is able to
reproduce the gravity driven microchannel flow of a gas bubble through liquid. In the regime of low
capillary
numbers
$\Ca$ (defined in Eq.~\ref{capillary:number:definition}), the flow is thoroughly described by
Bretherton \cite{bretherton},
with the deposition film thickness proportional to $\Ca^{2/3}$. However, there are deviations
from the Bretherton correlation for $\Ca\geq 0.005$, studied experimentally and numerically. This
work aims to examine the applicability of the lattice Boltzmann binary liquid model to simulations
of bubble flow in liquids for a range of $\Ca$ numbers. The binary
liquid model is a continuous-interface multiphase model allowing to properly resolve the interface.
It was found that the interface thickness needs to be $50$
percent of the film thickness for the simulation to give convergent results. When
the convergence criterion is fulfilled, the results are in a good agreement with the
numerical simulations of \citet{giavedoni-numerical} in the range of $\Ca$ from $0.01$ to $1$. 
\end{abstract}

@@

While there are a number
of works which simulate the Bretherton problem by the boundary value approach
\cite{ingham-plates,heil-bretherton}, those methods are of limited applicability for problems
involving complex geometries, free interface motion, or for problems involving coalescence and/or
droplet breakup. The continuous interface models are more flexible for those kind of simulations.
However, if the interface is spread over several grid nodes, the question of
proper film resolution in comparison with the interface resolution arises.  This work is focused on
resolving the parameter range of the binary liquid lattice
Boltzmann method to validate it for a range of capillary numbers.

The lattice Boltzmann method has emerged as a successful method to simulate
a wide variety of phenomena including hydrodynamics \cite{yu}, thermal flows
\cite{karlin-minimalmodels}, microflows \cite{ansumali-small-knudsen},
ferrofluids \cite{kuzmin-aniso}, and multiphase flows
\cite{swift,Shan-chen:extended}. Thanks to its kinetic nature, LBM as a particle
method easily tackles complex geometries and allows incorporation of
physical phenomena on the microscopic level, as in the case of multiphase models. Most
multiphase lattice Boltzmann models \cite{swift, Shan-chen:extended} resolve
the interface using continuous interface methods where the interfaces spans several grid nodes.
Such a representation
brings issues of the film thickness resolution versus interface
resolution -- the diffuse interface should be resolved in such a way as to have
negligible effect on the physics of the film.

@@

\section{Lattice Boltzmann benchmark}
To properly simulate and compare simulation results with the data
published in the literature, one needs to design a numerical benchmark addressing
the following challenges for the lattice Boltzmann method:
\begin{itemize}
 \item The diffuse nature of the interface in the multiphase model should not
influence the film thickness results. This is obviously achieved only for certain relation of the
interface resolution related to the film thickness. The same applies to spurious
currents arising due to surface tension discretization, which
can influence mass transfer in the thin film, thereby compromising the
overall solution. Therefore, the diffuse interface should not noticeably affect the film
thickness.
 \item The wettability coefficient which in principle can control the
dynamic angle and the menisci appearance \cite{pagonabarraga-finger} should not
influence the overall solution.
 \item The free-energy model examined in this work is a single density, two
viscosities model. The classical Bretherton problem is a free-interface problem between liquid and
gas
where the film thickness is established by itself. Thus, the viscosity and the density ratios for
the
classical
Bretherton problem should
be large. However, in terms of the lattice Boltzmann binary liquid framework only the viscosity
ratio can be addressed. Our computations show that the results are consistent with other works and
inertia effects can be neglected.
  \item The Bretherton problem addresses a bubble flow driven by a pressure
difference, not a body force. However, to the best authors' knowledge the pressure boundary
conditions are still not yet developed for the binary liquid model, because it involves the
treatment of associated gradients and Laplacians on the boundary.
This kind of pressure boundary conditions for the binary liquid model is currently
under development, and preliminary results show that the differences between flows driven by a body
force and by
pressure difference are small as far as the film
 thickness is of interest.  In this work we limit ourselves to the study
of body force driven flows, because of their simplicity and better numerical stability. This
case implies the periodic boundary conditions. As soon as the periodic boundary
conditions are applied, not a single bubble but whole bubble train is simulated. In this case one
needs
to ensure that the distance between bubbles is large enough to exclude mutual bubble influence.
 \item Most simulations are done for a channel with circular cross section,
which is quite difficult to address in terms of lattice Boltzmann free energy
framework, where one needs to calculate the Laplacian and gradients near boundaries.
Although one can apply certain finite difference stencils
\cite{arnold-boundary,hunt-boundary} developed specifically for curved boundaries,
the analysis of error introduced by the boundary conditions becomes an issue. Though the circular
axis-symmetric case can be simulated on rectangular grid by introducing specific mass and force
terms in the continuity and Navier-Stokes equation \cite{halliday-circular}, the lattice Boltzmann
incorporation of force and mass terms is in the process of development and validation. Therefore,
the benchmark is chosen as a plane-symmetric two-dimensional case.

@@

\subsection{The influence of the wall gradient}
The simulation results should not depend on the wall wettability. Wettability is defined
through the phase gradient \cite{pooley-contact}, $\partial_n \phi$.  We took a large
enough grid resolution in order for simulations to be consistent -- the grid size was
$177 \times 2626$ and the initial film thickness was
$12$ lattice Boltzmann units, which corresponds to the predicted film thickness.
We examined $11$ different values for the wall
gradient ranging from $-1$ to $1$. The parameters are summarized in Table
\ref{table:parameters:wall:gradient}. The results are consistent for all the
wall gradients given that the interface is properly resolved. For 
given wall gradients the values of interface thickness values are of $0.2\%$ relative accuracy.  For
the given wall gradients $0.8$
and $1.0$ the
simulations are unstable. One can see the calculated profiles in Fig.
\ref{fig:gradients:profiles}. The latter shows that the negative wall gradient values cause 
non-physical phase values (above $1.0$ near the wall). One should attribute them to the numerical
impose of wall gradients through the first-order finite difference scheme. Therefore, they do
not spoil numerical results. Note that
negative values of the phase gradient are
preferable, since the phase of the liquid adjacent to the wall has the value of $1$.
The phase values near the wall are above $1$ and do not interact
with the bubble which has phase value $-1$.  In case of positive
gradients the values near the wall are below $1$, and the gradient profile can
fuse with the values of the gas phase. This case corresponds to the slug flow when gas is in
contact with the wall.
\begin{figure}
\includegraphics[width=0.97\textwidth]{Figures/Wall/phase_grad_profiles.eps}
\caption{Phase profiles for different wall gradients. One can see that if the thickness
is properly resolved then the wettability of the wall does affect the film thickness. The
simulations were conducted for $200,000$ iterations. The measured profile was taken at the
nondimensional coordinate scaled to $H_{\mathrm{eff}}$ as $x=11.42$. $Ca=0.05$ was taken to
initialize simulation.
\label{fig:gradients:profiles}
}
\end{figure}

@@

\section{Conclusion}
This work presents numerical tips and a benchmark for the
Bretherton/Taylor problem using the binary liquid lattice Boltzmann method. To obtain correlations
we specifically designed the benchmark. The bubble was chosen sufficiently long for the film
thickness to
stabilize. Also, we used periodic boundary conditions for scheme robustness. Thus, the bubble
train was simulated instead of one bubble motion. To avoid the mutual influence of neighboring
bubbles a
distance between bubbles was chosen to be $3$ bubble sizes. The computational
results in terms of capillary number dependence and shape of a bubble show consistency with the
previously published data. Surprisingly, with large enough viscosity ratio the results are
independent on inertia. Also, we examined the influence of grid resolution on the results. We found
that if the phase interface is resolved as $50$ percent of the simulated film thickness, the
simulations
are convergent. Though  our results are specific to the binary liquid lattice
Boltzmann method, the numerical hints and procedures can be used for any
continuous interface method. 
\bibliographystyle{model1-num-names}
\bibliography{paper}
%\bibliographystyle{unsrtnat}
%\bibliography{paper}
\end{document}
